---
title: "HawkesModel"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package loading, include=FALSE}
library("hawkesbow")
library(readr)

```


## Fitting Hawkes model to marathon record data

Produce estimates for model, exponential Hawkes process, using MLE: 
- baseline intensity,
- reproduction mean  
- exponential fertility function rate

Note:  data is in days from the first world record which is set as time 0.  The 

```{r, warning = FALSE}
record_table_mod<-read_rds("record_table_mod.rds")

days_between = as.numeric(diff(record_table_mod$Date_ymd))
daysfromstart <- cumsum(days_between)
daysfromstart <- c(0,daysfromstart)  ### get data in terms of days from first record (time 0)

optMarathon<-mle(daysfromstart,"Exponential",40300)  # end date picked number greater than longest times
optMarathon$par
summary(optMarathon)
optMarathon$events
optMarathon$end



```

## Simulation using Hawkes model with MLE parameter estimates from data

Simulation for the same number of days used in the estimation.

```{r}
simRecs <- hawkes(40300, fun = optMarathon$par[1], repr = optMarathon$par[2], 
                  family = "exp", rate = optMarathon$par[3]) 
plot(simRecs, intensity = FALSE)
plot(simRecs, intensity = TRUE)
```


