---
title: "Is a sub 2 hour marathon in the near future?  Modeling rare events in sports."
author: Rodney X. Sturdivant, Ph.D., Baylor University and Nick Clark, Ph.D., West
  Point
output:
  beamer_presentation: default
  ioslides_presentation:
    logo: consulting_center_logo.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F)
```

```{r package loading, include=FALSE}
library(tidyverse);theme_set(theme_minimal()); theme_update(text = element_text(family = "serif"))
library(ggplot2)
library(MASS)
library(Hmisc)
library(DescTools)
library(goftest)
library(readr)
library("hawkesbow")
library(readr)
library(MASS)
```

# Outline

- Baseball Rare Events (if needed only)

- Background

- Marathon Data

- Simple Model

- Self-Exciting Model

- Further Research

# Background

## Are we living in a time of records?
- Include NY Times article screenshot of headline.
- Brief summary of article's premise.

## How can we address this question?
## What would randomness look like?

Pictures of Rod and Nick running

# Marathon World Record Data

Men's Marathon world records since 1908

NEED TO CLEAN UP - NICER TABLE WITH JUST TIME NAME NATIONALITY DATE
MAYBE INCLUDE A COUPLE OF PICTURES OF PEOPLE

```{r}
record_table_mod<-read_rds("record_table_mod.rds")
head(record_table_mod)
days_between = as.numeric(diff(record_table_mod$Date_ymd))
daysfromstart <- cumsum(days_between)
daysfromstart <- c(0,daysfromstart)  ### get data in terms of days from first record (time 0)
  # units of year
daysfromstart_mod2 <- daysfromstart/365 
days_between_mod2 = diff(daysfromstart_mod2)

```

# Visualizing the data

AGAIN NEED CLEANING UP - NEED TIMES IN SOMETHING OTHER THAN SECONDS MAYBE INCLUDE 2 HOUR HORIZONTAL BAR
MAYBE HAVE A SLIDE WHERE ADD PICTURES OF PEOPLE WHO LOWERED RECORD BY A LOT
WHICH PLOT(S) TO USE?  MAYBE TWO OF THEM BUT ON ONE SLIDE?

```{r}


ggplot() + 
  geom_segment(data=record_table_mod, aes(x=Date_ymd,xend=end,y=Time_sec,yend=Time_sec),size=2) + 
  ylab("Time in Seconds")+
  xlab("Marathon Record")+
  geom_rug(data=record_table_mod,aes(x=Date_ymd,y=Time_sec,),sides="b")
```

# Visualize B
```{r}
p1<-plot(record_table_mod$Date_ymd, record_table_mod$Time_t, type = "h",
     xlab = "Data (YMD)", ylab = "Record in Seconds")

```



# Visualize C

```{r}

record_table_mod %>% ggplot() +
  geom_line(aes(x=Date_ymd, y=Time_sec, color = "Record in Seconds"),size=2) +
  scale_color_manual(name="",
                     values = c("Record in Seconds"="#00ba38")) +
  theme(panel.grid.minor = element_blank())


```

# SIMPLE MODEL

## POISSON PROCESS (NEED TO SHORTEN, OR PUT ON TWO SLIDES)

A model for a series of discrete events where the average time between events is known, but the exact timing of events is "random" meeting the following criteria: 

-  Events are independent of each other. The occurrence of one event does not affect the probability another event will occur.

-  The average rate (events per time period) is constant.

-  Two events cannot occur at the same time.

The time between events (known as the interarrival times) follow an exponential distribution defined as:

$$P(T>t) = e^{-\lambda t}$$
Where T is the random variable of the time until the next event, t is a specific time for the next event, and $\lambda$ is the rate:  the average number of events per unit of time.  Note the possible values of T are greater than 0 (positive only).  


# Reasonableness of Exponential Interarrivals

The exponential distribution has certain attributes, for example:

$E(T) = 1/\lambda$
$SD(T) = 1/\lambda$

The mean and standard deviation of the years between records:

```{r}

mean(days_between_mod2)
sd(days_between_mod2)

hist(days_between_mod2)
```

#  MORE ON THE SIMPLE MODEL

We estimate (MLE) 
$\lambda = 1/E(T)$

```{r}
expfit=fitdistr(days_between_mod2,"exponential")
exprate<-expfit$estimate
exprate
```
Model fit  

```{r}
x=density(days_between_mod2)
hist(days_between_mod2,main="Histogram, density curve and exponential model",xlab="Interarrival Time",freq=FALSE,ylim=c(0,0.5))
lines(x,col="red",lty = 2, lwd = 2)
curve(dexp(x, rate = exprate),col=3, lty = 3, lwd = 2, add = TRUE)

```

# Fit of simple model

```{r}
ks.test(days_between_mod2,pexp,rate=exprate)

cvm.test(days_between_mod2,"pexp",rate=exprate,estimated = TRUE)
ad.test(days_between_mod2,"pexp",rate=exprate,estimated = TRUE)


```

# Fit of simple model B



```{r}
x=seq(0,max(days_between_mod2),0.1)
plot(x,pexp(x,rate=exprate),type="l",col="red", main="EDF and Exponential CDF",xlab="Interarrival Time",ylab="Proportion <= x")
Ecdf(days_between_mod2,xlab='Interarrival Times',subtitle=FALSE,add=TRUE)
```


# Are records then random?

```{r}

PlotQQ(days_between_mod2, function(p) qexp(p, rate=exprate))

```

# What are the poorly fit points?

LOOK BACK AT THE ORIGINAL DATA HERE...LONGEST TIMES BETWEEN EVENTS (I THINK)...ONE IS WW2 PRETTY SURE...THE OTHER NEED TO LOOK AGAIN - MAYBE AN UNUSUALLY LARGE LOWERING OF THE RECORD OR SOMETHING?

# References

Data source: Wikipedia (https://en.wikipedia.org/wiki/Marathon_world_record_progression) scraped August 12, 2022

Poisson process:
https://towardsdatascience.com/the-poisson-distribution-and-poisson-process-explained-4e2cb17d459



